---
title: "Chapter 12"
author: "Scott Spencer"
date: "8/30/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, message = FALSE, error = FALSE)
library(dplyr); library(tidyr); library(rstan); library(skimr); library(ggplot2); library(ggthemes)
theme_set(theme_tufte(base_family = 'sans'))
```

The code below is meant as a directly-in-Stan translation of the examples in Chapter 12 of McElreath's *Statistical Rethinking*.

## 12.1. Example: Multilevel tadpoles

Load the data.

```{r}
data('reedfrogs', package = 'rethinking')
d <-  reedfrogs; rm(reedfrogs)
d <- d %>% mutate(tank = seq(NROW(d)))
```

Code first model in Stan.

```{stan output.var="m12_1"}
data {
  int N;
  int N_tanks;
  int tank_id[N];
  int density[N];
  int surv[N];
}
parameters {
  vector[N_tanks] a_tank;
}
model {
  vector[N] p;
  target += normal_lpdf(a_tank | 0, 5);
  for(i in 1:N) p[i] = inv_logit( a_tank[tank_id[i]] );
  target += binomial_lpmf(surv | density, p);
}
generated quantities {
  vector[N] log_lik;
  {
  vector[N] p;
  for(i in 1:N) {
    p[i] = inv_logit( a_tank[tank_id[i]] );
    log_lik[i] = binomial_lpmf(surv[i] | density[i], p[i]);
  }
  }
}

```

Organize data and sample from model.

```{r}
dat <- list(
  N = NROW(d),
  density = d$density,
  N_tanks = max(d$tank),
  tank_id = d$tank,
  surv = d$surv
)

fit12_1 <- sampling(m12_1, data = dat, iter = 1000, chains = 2, cores = 2)
```

Summary

```{r}
print(fit12_1, include = F, pars = "log_lik", probs = c(.1, .5, .9))
```

Code second, multilevel in Stan.

```{stan output.var="m12_2"}
data {
  int N;
  int N_tanks;
  int tank_id[N];
  int density[N];
  int surv[N];
}
parameters {
  real alpha;
  real<lower=0> sigma;
  vector[N_tanks] a_tank;
}
model {
  vector[N] p;
  target += normal_lpdf(alpha | 0, 1);
  target += cauchy_lpdf(sigma | 0, 1);
  target += normal_lpdf(a_tank | alpha, sigma);
  for(i in 1:N) p[i] = inv_logit( a_tank[tank_id[i]] );
  target += binomial_lpmf(surv | density, p);
}
generated quantities {
  vector[N] log_lik;
  {
  vector[N] p;
  for(i in 1:N) {
    p[i] = inv_logit( a_tank[tank_id[i]] );
    log_lik[i] = binomial_lpmf(surv[i] | density[i], p[i]);
  }
  }
}

```

Sample from model.

```{r}
fit12_2 <- sampling(m12_2, data = dat, iter = 1000, chains = 2, cores = 2)
```

Summary

```{r}
print(fit12_2, include = F, pars = "log_lik", probs = c(.1, .5, .9))
```

Compare models.

```{r}
library(loo)
ll12_1 <- extract_log_lik(fit12_1)
ll12_2 <- extract_log_lik(fit12_2)
reff12_1 <- relative_eff(ll12_1, chain_id = c(rep(1, 500), rep(2, 500)), cores =2)
reff12_2 <- relative_eff(ll12_2, chain_id = c(rep(1, 500), rep(2, 500)), cores =2)

waic12_1 <- waic(ll12_1, r_eff = reff12_1, cores = 2)
waic12_2 <- waic(ll12_2, r_eff = reff12_2, cores = 2)
loo::compare(waic12_1, waic12_2)

loo12_1 <- loo(ll12_1, r_eff = reff12_1, cores = 2)
loo12_2 <- loo(ll12_2, r_eff = reff12_1, cores = 2)
loo::compare(loo12_1, loo12_2)
```

Review estimates for each tank.

Figure 12.1

```{r}
post12_2 <- as.data.frame(fit12_2, pars = c('alpha', 'sigma', 'a_tank'))
d$propsurv_est <- post12_2[,-c(1:2)] %>% colMeans %>% plogis

ggplot(d) + 
  geom_vline(xintercept = c(16.5, 32.5)) +
  geom_hline(yintercept = plogis(mean(post12_2$alpha)), linetype = 'dashed') +
  geom_point(aes(x = tank, y = propsurv), color = 'dodgerblue') +
  geom_point(aes(x = tank, y = propsurv_est), shape = 21) +
  scale_x_continuous(limits = c(0, 48), breaks = c(1, 16, 32, 48)) +
  annotate('text', x = 8, y = 0, label = 'Small tanks', hjust = 0.5) +
  annotate('text', x = 16+8, y = 0, label = 'Medium tanks', hjust = 0.5) +
  annotate('text', x = 32+8, y = 0, label = 'Large tanks', hjust = 0.5) +
  labs(x = 'Tank', y = 'Proportion survived')

```

Simulate survival rates.

Figure 12.2 left side

```{r}
for(i in 1:100) 
  curve(dnorm(x, post12_2$alpha[i], post12_2$sigma[i]), 
        add = T, 
        col = alpha('black', .2), 
        xlim = c(-3, 4), 
        ylim = c(0, .35), 
        xlab = 'log-odds survival', 
        ylab = 'Density')
```

Figure 12.2 right side

```{r}
sim_tanks <- 
  rnorm(8000, post12_2$alpha, post12_2$sigma) %>% 
  plogis

ggplot() + 
  geom_density(aes(sim_tanks), fill = 'skyblue') +
  labs(x = "Probability survival", y = "Density")
```

## 12.2 Varying effects and the underfitting/overfitting trade-off

### 12.2.2. Assign values to the parameters.

```{r}
a <- 1.4
sigma <- 1.5
nponds <- 60
ni <- as.integer( rep( c(5, 15, 25, 35), each = 15))
a_pond <- rnorm(nponds, mean = a, sd = sigma)
dsim <- data.frame(pond = 1:nponds, ni=ni, true_a = a_pond)
```


### 12.2.3 Simulate survivors

```{r}
dsim$si <- rbinom(nponds, prob = plogis(dsim$true_a), size = dsim$ni)
```

### 12.2.4 compute the no-pooling estimates

```{r}
dsim$no_pool <- dsim$si / dsim$ni
```

### 12.2.5 compute the partial-pooling estimates

```{stan output.var="m12_3"}
data {
  int N;
  int N_ponds;
  int pond[N];
  int si[N];
  int ni[N];
}
parameters {
  real a;
  real<lower=0> sigma;
  vector[N_ponds] a_pond;
}
model {
  vector[N] p;
  target += normal_lpdf(a | 0, 1);
  target += cauchy_lpdf(sigma | 0, 1);
  target += normal_lpdf(a_pond | a, sigma);
  for(i in 1:N) p[i] = inv_logit(a_pond[pond[i]]);
  target += binomial_lpmf(si | ni, p);
}

```

Organize data and sample from model.

```{r}
dat <- list(
  N = NROW(dsim),
  N_ponds = max(dsim$pond),
  pond = dsim$pond,
  si = dsim$si,
  ni = dsim$ni
)

fit12_3 <- sampling(m12_3, data = dat, iter = 1000, chains = 2, cores = 2)
```

Summarise model.

```{r}
print(fit12_3, probs = c(.1, .5, .9))
```

Figure 12.3

```{r}
post12_3 <- as.data.frame(fit12_3, pars = 'a_pond')
dsim$partpool <- plogis(colMeans(post12_3))
dsim <- dsim %>% 
  mutate(p_true = plogis(true_a),
         nopool_err = abs(no_pool - p_true),
         partpool_err = abs(partpool - p_true)) %>%
  group_by(ni) %>% 
  mutate(np_err_mean = mean(nopool_err),
         pp_err_mean = mean(partpool_err)) %>% 
  ungroup

ggplot(dsim) + 
  geom_point(aes(x = pond, y = nopool_err), color = 'dodgerblue') +
  geom_point(aes(x = pond, y = partpool_err), shape = 21) +
  geom_line(aes(x = pond, y = np_err_mean, group = ni), color = 'dodgerblue') +
  geom_line(aes(x = pond, y = pp_err_mean, group = ni), linetype = 'dashed') +
  geom_vline(xintercept = 15 * seq(3) + .5) +
  annotate('text', x = 8, y = 0.4, label = 'tiny (5)') +
  annotate('text', x = 23, y = 0.4, label = 'small (15)') +
  annotate('text', x = 38, y = 0.4, label = 'medium (25)') +
  annotate('text', x = 53, y = 0.4, label = 'large (35)') +
  labs(x = 'pond', y = 'absolute error')
  
```

## 12.3 More than one type of cluster

### 12.3.1 multilevel chimpanzees

Code the model

```{stan output.var="m12_4"}
data {
  int N;
  int C[N];
  int P[N];
  int N_actors;
  int actor[N];
  int L[N];
}
parameters {
  real a;
  real bp;
  real bpc;
  real<lower=0> sigma_actor;
  vector[N_actors] a_actor;
}
model {
  vector[N] p;
  for (i in 1:N) p[i] = a + a_actor[actor[i]] + (bp + bpc * C[i]) * P[i];

  target += normal_lpdf( a | 0, 10);
  target += normal_lpdf( bp | 0, 10);
  target += normal_lpdf( bpc | 0, 10);
  target += normal_lpdf( a_actor | 0, sigma_actor);
  target += cauchy_lpdf(sigma_actor | 0, 1);

  target += binomial_logit_lpmf(L | 1, p);
}
generated quantities {
  vector[N] log_lik;
  {
  vector[N] p;
  for (i in 1:N) {
    p[i] = a + a_actor[actor[i]] + (bp + bpc * C[i]) * P[i];
    log_lik[i] = binomial_logit_lpmf(L[i] | 1, p[i]);
  }
  }
}

```

Organize data and sample from model.

```{r}
data('chimpanzees', package = 'rethinking')
d <- chimpanzees; rm(chimpanzees)
dat <- list(
  N = NROW(d),
  C = d$condition,
  P = d$prosoc_left,
  N_actors = max(d$actor),
  actor = d$actor,
  L = d$pulled_left
)

fit12_4 <- sampling(m12_4, data = dat, iter = 4000, chains = 4, cores = 4)
```

Model summary

```{r}
print(fit12_4, include = F, pars = 'log_lik', probs = c(.1, .5, .9))
```


```{r}
post12_4 <- as.data.frame(fit12_4)
total_a_actor <- sweep(post12_4[,5:11], 2, FUN = "+", post12_4$a) %>% colMeans
print(total_a_actor, digits = 2)
```

### 12.3.2 Two types of cluster

```{stan output.var="m12_5"}
data {
  int N;
  int N_actors;
  int N_blocks;
  int actors[N];
  int blocks[N];
  int C[N];
  int P[N];
  int L[N];
}
parameters {
  real a;
  real bp;
  real bpc;
  vector[N_actors] a_actor;
  vector[N_blocks] a_block;
  real<lower=0> sigma_actor;
  real<lower=0> sigma_block;
}
model {
  vector[N] p;
  target += normal_lpdf( a | 0, 10);
  target += normal_lpdf( bp | 0, 10);
  target += normal_lpdf( bpc | 0, 10);
  target += normal_lpdf( a_actor | 0, sigma_actor);
  target += normal_lpdf( a_block | 0, sigma_block);
  target += cauchy_lpdf( sigma_actor | 0, 5);
  target += cauchy_lpdf( sigma_block | 0, 5);

  for(i in 1:N) p[i] = a + a_actor[actors[i]] + a_block[blocks[i]] + (bp + bpc * C[i]) * P[i];
  target += binomial_logit_lpmf(L | 1, p);
}
generated quantities {
  vector[N] log_lik;
  {
  vector[N] p;
  for(i in 1:N) {
    p[i] = a + a_actor[actors[i]] + a_block[blocks[i]] + (bp + bpc * C[i]) * P[i];
    log_lik[i] = binomial_logit_lpmf(L[i] | 1, p[i]);  

  }
  }
}

```

Organize data and sample from model.

```{r}
dat <- list(
  N = NROW(d),
  C = d$condition,
  P = d$prosoc_left,
  N_actors = max(d$actor),
  N_blocks = max(d$block),
  actors = d$actor,
  blocks = d$block,
  L = d$pulled_left
)

fit12_5 <- sampling(m12_5, data = dat, iter = 4000, chains = 4, cores = 4)
```

Model summary

```{r}
print(fit12_5, include = F, pars = 'log_lik', probs = c(.1, .5, .9))
```

Look at the sigmas.

```{r}
post12_5 <- as.data.frame(fit12_5)
ggplot() + theme_tufte(base_family = 'sans') +
  geom_density(aes(post12_5$sigma_actor)) +
  geom_density(aes(post12_5$sigma_block))
```

Compare models

```{r}
ll12_4 <- extract_log_lik(fit12_4)
ll12_5 <- extract_log_lik(fit12_5)

reff12_4 <- relative_eff(ll12_4, chain_id = c(rep(1, 2000), rep(2, 2000), rep(3, 2000), rep(4, 2000)), cores =4)
reff12_5 <- relative_eff(ll12_5, chain_id = c(rep(1, 2000), rep(2, 2000), rep(3, 2000), rep(4, 2000)), cores =4)

waic12_4 <- waic(ll12_4, r_eff = reff12_4, cores = 2)
waic12_5 <- waic(ll12_5, r_eff = reff12_5, cores = 2)
loo::compare(waic12_4, waic12_5)

loo12_4 <- loo(ll12_4, r_eff = reff12_4, cores = 2)
loo12_5 <- loo(ll12_5, r_eff = reff12_5, cores = 2)
loo::compare(loo12_4, loo12_5)
```

## 12.4 Multilevel posterior predictions

### 12.4.1 posterior prediction for same clusters

```{r}
nd <- data.frame(actor = 1, 
                 P = c(0, 1, 0, 1), 
                 C = c(0, 0, 1, 1))

f_mu <- function(actor, C, P, new = F) {
  logodds <- matrix()
  if( new == F)
    logodds <- post12_4[,'a'] + post12_4[,4+actor] + (post12_4[,'bp'] + post12_4[,'bpc'] * C) * P
  else logodds <- post12_4[,'a'] + (post12_4[,'bp'] + post12_4[,'bpc'] * C) * P
  return(logodds)
}

logodds <- mapply(f_mu, actor =nd$actor, C = nd$C, P = nd$P)
p_mu <- plogis(logodds)
p_mu_mean <- apply(p_mu, 2, mean)
p_mu_hpdi <- apply(p_mu, 2, rethinking::HPDI)

nd <- nd %>% mutate(PC = interaction(P, C, sep = '/'),
                    p_mu = p_mu_mean,
                    p_mu_hpdi_l = p_mu_hpdi[1,],
                    p_mu_hpdi_h = p_mu_hpdi[2,])

ggplot(nd) + 
  geom_ribbon(aes(x = 1:4, 
                  ymin = p_mu_hpdi_l,
                  ymax = p_mu_hpdi_h),
              alpha = .1) +
  geom_line(aes(x = 1:4, 
                y = p_mu)) +
  scale_x_continuous(breaks = 1:4, labels = nd$PC) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = 'Prosocial Left/Condition', y = 'Proportion pulled left', subtitle = "Actor 1")

```

### 12.4.2 posterior prediction with new clusters

Review average actor.

```{r}
nd <- data.frame(actor = 0, 
                 P = c(0, 1, 0, 1), 
                 C = c(0, 0, 1, 1))

logodds <- mapply(f_mu, actor =nd$actor, C = nd$C, P = nd$P, new = T)    
p_mu <- plogis(logodds)
p_mu_mean <- apply(p_mu, 2, mean)
p_mu_hpdi <- apply(p_mu, 2, rethinking::PI, prob = .8)

nd <- nd %>% mutate(PC = interaction(P, C, sep = '/'),
                    p_mu = p_mu_mean,
                    p_mu_hpdi_l = p_mu_hpdi[1,],
                    p_mu_hpdi_h = p_mu_hpdi[2,])

p1 <- ggplot(nd) +
  geom_ribbon(aes(x = 1:4, 
                  ymin = p_mu_hpdi_l,
                  ymax = p_mu_hpdi_h),
              alpha = .1) +
  geom_line(aes(x = 1:4, 
                y = p_mu)) +
  scale_x_continuous(breaks = 1:4, labels = nd$PC) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = 'Prosocial Left/Condition', y = 'Proportion pulled left', subtitle = 'Average Actor')

```

Get marginal of actors.

```{r}
a_actor_sims <- rnorm(8000, 0, post12_4$sigma_actor)
sim_logodds <- logodds + a_actor_sims
sim_mu <- plogis(sim_logodds)

sim_mu_mean <- apply(sim_mu, 2, mean)
sim_mu_hpdi <- apply(sim_mu, 2, rethinking::PI, prob = .8)

nd <- nd %>% mutate(sim_mu = sim_mu_mean,
                    sim_mu_hpdi_l = sim_mu_hpdi[1,],
                    sim_mu_hpdi_h = sim_mu_hpdi[2,])

p2 <- ggplot(nd) + 
  geom_ribbon(aes(x = 1:4, 
                  ymin = sim_mu_hpdi_l,
                  ymax = sim_mu_hpdi_h),
              alpha = .1) +
  geom_line(aes(x = 1:4, 
                y = sim_mu)) +
  scale_x_continuous(breaks = 1:4, labels = nd$PC) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = 'Prosocial Left/Condition', y = 'Proportion pulled left', subtitle = 'Marginal of Actor')

```

Show 50 simulated actors.

```{r}
p3 <- ggplot() + theme_tufte(base_family = 'sans') +
  geom_line(aes(x = rep(1:4, 50), 
                y = as.vector(t(sim_mu[1:50,])), 
                group = rep(1:50, each = 4))) + 
  labs(x = 'Prosocial Left / Condition', y = 'Proportion pulled left', subtitle = '50 Simulations of actor')
```

Figure 12.5

```{r, fig.width=8}
library(gridExtra)
grid.arrange(p1, p2, p3, nrow = 1)
```

### 12.4.3 focus and multilevel prediction

Load data.

```{r}
data('Kline', package = 'rethinking')
d <- Kline; rm(Kline)
```

Code model in Stan.

```{stan output.var="m12_6"}
data {
  int N;
  int T[N];
  int N_societies;
  int society[N];
  int P[N];
}
parameters {
  real alpha;
  vector[N_societies] a_society;
  real bp;
  real<lower=0> sigma_society;
}
model {
  vector[N] mu;
  target += normal_lpdf(alpha | 0, 10);
  target += normal_lpdf(bp | 0, 1);
  target += cauchy_lpdf(sigma_society | 0, 1);
  target += normal_lpdf(a_society | 0, sigma_society);
  for(i in 1:N) mu[i] = alpha + a_society[society[i]] + bp * log(P[i]);
  target += poisson_log_lpmf(T | mu);
}
generated quantities {
  vector[N] log_lik;
  {
  vector[N] mu;
  for(i in 1:N) {
    mu[i] = alpha + a_society[society[i]] + bp * log(P[i]);
    log_lik[i] = poisson_log_lpmf(T[i] | mu[i]);
  }
  }
}

```

Organize data and sample from model.

```{r}
dat <- list(
  N = NROW(d),
  T = d$total_tools,
  N_societies = 10,
  society = 1:10,
  P = d$population
)

fit12_6 <- sampling(m12_6, data = dat, iter = 1000, chains = 2, cores = 2)
```

Model summary

```{r}
print(fit12_6, include = F, pars = 'log_lik', probs = c(.1, .5, .9))
```

Look at effective parameters.

```{r}
ll12_6 <- extract_log_lik(fit12_6)
reff12_6 <- relative_eff(ll12_6, chain_id = c(rep(1, 500), rep(2, 500)), cores =2)
waic12_6 <- waic(ll12_6, r_eff = reff12_6, cores = 2)
waic12_6
```

Posterior predictions to visualize the overdispersion

```{r}
nd <- data.frame(
  P = exp(seq(6, 14, length.out = 30)),
  society = rep(1, 30)
  )

post12_6 <- as.data.frame(fit12_6, include = F, pars = 'log_lik')

f_mu <- function(P) post12_6[, 'alpha'] + post12_6[, 'bp'] * log(P)
sim_mu <- mapply(f_mu, P = nd$P)
sim_sigma <- rnorm(1000, 0, post12_6$sigma_society)
sim <- sim_mu + sim_sigma
sim_mean <- apply(sim, 2, mean)
sim_hpdi97 <- apply(sim, 2, rethinking::PI, prob = .97)
sim_hpdi89 <- apply(sim, 2, rethinking::PI, prob = .89)
sim_hpdi67 <- apply(sim, 2, rethinking::PI, prob = .67)
nd <- nd %>% mutate(sim_mean = sim_mean,
                    sim_hpdi_l97 = sim_hpdi97[1,],
                    sim_hpdi_h97 = sim_hpdi97[2,],
                    sim_hpdi_l89 = sim_hpdi89[1,],
                    sim_hpdi_h89 = sim_hpdi89[2,],
                    sim_hpdi_l67 = sim_hpdi67[1,],
                    sim_hpdi_h67 = sim_hpdi67[2,])
```

Figure 12.6

```{r}
ggplot(nd) + 
  geom_ribbon(aes(x = log(P), ymin = exp(sim_hpdi_l97), ymax = exp(sim_hpdi_h97)), alpha = .1) +
  geom_ribbon(aes(x = log(P), ymin = exp(sim_hpdi_l89), ymax = exp(sim_hpdi_h89)), alpha = .1) +
  geom_ribbon(aes(x = log(P), ymin = exp(sim_hpdi_l67), ymax = exp(sim_hpdi_h67)), alpha = .1) +
  geom_line(aes(x = log(P), y = exp(sim_mean))) +
  theme(legend.position = "") + 
  coord_fixed(ratio = .1, xlim = c(6, 13), ylim = c(0, 70)) +
  scale_x_continuous(breaks = seq(7, 12, by = 1)) + 
  scale_y_continuous(breaks = seq(0, 70, by = 10)) +
  labs(x = 'log population', y = 'total tools')
```




