---
title: "Chapter 07"
author: "Scott Spencer"
date: "8/27/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, message = FALSE, error = FALSE)
library(dplyr); library(tidyr); library(rstan); library(ggplot2); library(ggthemes)
theme_set(theme_tufte(base_family = 'sans'))
```

The code below is meant as a directly-in-Stan translation of the examples in Chapter 7 of McElreath's *Statistical Rethinking*.

## 7.1 Building an interaction

Load data.

```{r}
data('rugged', package = 'rethinking')
d  <- rugged; rm(rugged)
```

Transform data.

```{r}
d$loggdp <- log(d$rgdppc_2000)
d <- d[complete.cases(d$loggdp),]
d.A1 <- d[d$cont_africa==1,]
d.A0 <- d[d$cont_africa==0,]
```

Code model.

```{stan output.var="m07_1"}
data {
  int N;
  vector[N] loggdp;
  vector[N] rugged;
}
parameters {
  real a;
  real bR;
  real<lower=0, upper=10> sigma;
}
model {
  vector[N] mu = a + bR * rugged;
  target += normal_lpdf(loggdp | mu, sigma);
  target += normal_lpdf(a | 8, 100);
  target += normal_lpdf(bR | 0, 1);
}

```

Organize data and sample from model.

```{r}
m07_2 <- m07_1

dat <- list(
  N = NROW(d.A1),
  loggdp = d.A1$loggdp,
  rugged = d.A1$rugged
)

fit07_1 <- sampling(m07_1, data = dat, iter = 10000, chains = 4, cores = 4)
```


```{r}
post <- as.data.frame(fit07_1)

mu <- post$a + d.A1$rugged %*% t(post$bR)
mu_mean <- rowMeans(mu)
mu_hpdi <- apply(mu, 1, rethinking::HPDI)

# simulate loggdp
iter <- 1e5
y_hat <- matrix(nrow = iter, ncol = NROW(d.A1))
for(i in 1:NROW(d.A1)) y_hat[,i] <- rnorm(iter, post[,"a"] + d.A1[i,"rugged"] %*% t(post[,"bR"]), post[,"sigma"])

# get stats on sim
y_hat_avg <- colMeans(y_hat)
y_hat_pi <- apply(y_hat, 2, rethinking::PI)
```

Figure 7.2a

```{r}
ggplot() + 
  geom_point(data = d.A1,
             aes(rugged, loggdp), shape = 21, color = 'dodgerblue') +
  #geom_ribbon(aes(x = d.A1$rugged, ymin = y_hat_pi[1,], ymax = y_hat_pi[2,]), alpha = .2) +
  geom_ribbon(aes(x = d.A1$rugged, ymin = mu_hpdi[1,], ymax = mu_hpdi[2,]), alpha = .2) +
  geom_abline(slope = mean(post$bR), intercept = mean(post$a) )
```

```{r}
dat <- list(
  N = NROW(d.A0),
  loggdp = d.A0$loggdp,
  rugged = d.A0$rugged
)

fit07_2 <- sampling(m07_2, data = dat, iter = 10000, chains = 4, cores = 4)
```


```{r}
post <- as.data.frame(fit07_2)

mu <- post$a + d.A0$rugged %*% t(post$bR)
mu_mean <- rowMeans(mu)
mu_hpdi <- apply(mu, 1, rethinking::HPDI)

# simulate loggdp
iter <- 1e5
y_hat <- matrix(nrow = iter, ncol = NROW(d.A0))
for(i in 1:NROW(d.A0)) y_hat[,i] <- rnorm(iter, post[,"a"] + d.A0[i,"rugged"] %*% t(post[,"bR"]), post[,"sigma"])

# get stats on sim
y_hat_avg <- colMeans(y_hat)
y_hat_pi <- apply(y_hat, 2, rethinking::PI)
```

Figure 7.2b

```{r}
ggplot() + 
  geom_point(data = d.A0,
             aes(rugged, loggdp), shape = 21, color = 'dodgerblue') +
  #geom_ribbon(aes(x = d.A0$rugged, ymin = y_hat_pi[1,], ymax = y_hat_pi[2,]), alpha = .2) +
  geom_ribbon(aes(x = d.A0$rugged, ymin = mu_hpdi[1,], ymax = mu_hpdi[2,]), alpha = .2) +
  geom_abline(slope = mean(post$bR), intercept = mean(post$a) )
```

### 7.1.1 Adding a dummy variable doesn't work

First re-sample model above with all data (with and without Africa). Need to re-code model to include log likelihood in `generated quantities` block since we will be comparing the two models.

```{stan output.var="m07_3"}
data {
  int N;
  vector[N] loggdp;
  vector[N] rugged;
}
parameters {
  real a;
  real bR;
  real<lower=0, upper=10> sigma;
}
model {
  vector[N] mu = a + bR * rugged;
  target += normal_lpdf(loggdp | mu, sigma);
  target += normal_lpdf(a | 8, 100);
  target += normal_lpdf(bR | 0, 1);
}
generated quantities {
  vector[N] log_lik;
  {
  vector[N] mu;
  for(n in 1:N) {
    mu[n] = a + bR * rugged[n];
    log_lik[n] = normal_lpdf(loggdp[n] | mu[n], sigma);
  }
  }
}

```

```{r}
dat <- list(
  N = NROW(d),
  loggdp = d$loggdp,
  rugged = d$rugged
)

fit07_3 <- sampling(m07_3, data = dat, iter = 10000, chains = 4, cores = 4)
```

Now code a model that uses a dummary variable for Africa.

```{stan output.var="m07_4"}
data {
  int N;
  vector[N] loggdp;
  vector[N] rugged;
  vector<lower=0,upper=1>[N] cont_africa;
}
parameters {
  real a;
  real bR;
  real bA;
  real<lower=0, upper=10> sigma;
}
model {
  vector[N] mu = a + bR * rugged + bA * cont_africa;
  target += normal_lpdf(loggdp | mu, sigma);
  target += normal_lpdf(a | 8, 100);
  target += normal_lpdf(bR | 0, 1);
  target += normal_lpdf(bA | 0, 1);
}
generated quantities {
  vector[N] log_lik;
  {
  vector[N] mu;
  for(n in 1:N) {
    mu[n] = a + bR * rugged[n] + bA * cont_africa[n];
    log_lik[n] = normal_lpdf(loggdp[n] | mu[n], sigma);
  }
  }
}

```

```{r}
dat <- list(
  N = NROW(d),
  loggdp = d$loggdp,
  rugged = d$rugged,
  cont_africa = d$cont_africa
)

fit07_4 <- sampling(m07_4, data = dat, iter = 10000, chains = 4, cores = 4)
```

Calculate WAIC and compare models

```{r}
# compare models
library(loo)

log_lik_07_3 <- extract_log_lik(fit07_3, merge_chains = FALSE)
waic07_3 <- waic(log_lik_07_3)

log_lik_07_4 <- extract_log_lik(fit07_4, merge_chains = FALSE)
waic07_4 <- waic(log_lik_07_4)

mod_comp <- compare(waic07_3, waic07_4)
mod_comp
```

TODO: above does not include model weighting. 

Plot predictions using model `m07_4`.

```{r}
post <- as.data.frame(fit07_4)

mu <- post$a + d$rugged %*% t(post$bR) + d$cont_africa %*% t(post$bA)
mu_mean <- rowMeans(mu)
mu_hpdi <- apply(mu, 1, rethinking::HPDI)

# simulate loggdp
iter <- 1e5
y_hat <- matrix(nrow = iter, ncol = NROW(d))
for(i in 1:NROW(d)) y_hat[,i] <- rnorm(iter, post[,"a"] + d[i,"rugged"] %*% t(post[,"bR"]) + d[i,"cont_africa"] %*% t(post[,"bA"]), post[,"sigma"])

# get stats on sim
y_hat_avg <- colMeans(y_hat)
y_hat_pi <- apply(y_hat, 2, rethinking::PI)
```

Figure 7.3

```{r}
d_04 <- d %>% 
  mutate(mu_mean = mu_mean,
         mu_hpdi_l = mu_hpdi[1,],
         mu_hpdi_h = mu_hpdi[2,])

ggplot(d_04) +
  geom_point(aes(rugged, loggdp, color = cont_africa), shape = 21) +
  theme(legend.position = '') +
  geom_line(data = d_04 %>% filter(cont_africa == 0),
            aes(rugged, mu_mean)) +
  geom_ribbon(data = d_04 %>% filter(cont_africa == 0),
              aes(x=rugged, ymin=mu_hpdi_l, ymax=mu_hpdi_h), alpha = .1) +
  geom_line(data = d_04 %>% filter(cont_africa == 1),
            aes(rugged, mu_mean), color = 'dodgerblue') +
  geom_ribbon(data = d_04 %>% filter(cont_africa == 1),
              aes(x=rugged, ymin=mu_hpdi_l, ymax=mu_hpdi_h), alpha = .1, fill = 'dodgerblue') +
  annotate('text', x = 4, y = 9, label = "Not Africa") +
  annotate('text', x = 4, y = 7, label = "Africa", color = 'dodgerblue') 
```

### 7.1.2 Adding a linear interaction does work

Code the model.

```{stan output.var="m07_5"}
data {
  int N;
  vector[N] loggdp;
  vector[N] rugged;
  vector<lower=0,upper=1>[N] cont_africa;
}
parameters {
  real a;
  real bR;
  real bA;
  real bAR;
  real<lower=0, upper=10> sigma;
}
model {
  vector[N] gamma;
  vector[N] mu;
  
  for (i in 1:N) {
    gamma[i] = bR + bAR * cont_africa[i];
    mu[i] = a + gamma[i] * rugged[i] + bA * cont_africa[i];
  }

  target += normal_lpdf(loggdp | mu, sigma);
  target += normal_lpdf(a | 8, 100);
  target += normal_lpdf(bR | 0, 1);
  target += normal_lpdf(bA | 0, 1);
  target += normal_lpdf(bAR | 0, 1);
}
generated quantities {
  vector[N] log_lik;
  {
  vector[N] mu; vector[N] gamma;
  for(n in 1:N) {
    gamma[n] = bR + bAR * cont_africa[n];
    mu[n] = a + gamma[n] * rugged[n] + bA * cont_africa[n];
    log_lik[n] = normal_lpdf(loggdp[n] | mu[n], sigma);
  }
  }
}

```

```{r}
dat <- list(
  N = NROW(d),
  loggdp = d$loggdp,
  rugged = d$rugged,
  cont_africa = d$cont_africa
)

fit07_5 <- sampling(m07_5, data = dat, iter = 10000, chains = 4, cores = 4)
```

Compare three models.

```{r}
log_lik_07_5 <- extract_log_lik(fit07_5, merge_chains = FALSE)
waic07_5 <- waic(log_lik_07_5)

mod_comp <- compare(waic07_3, waic07_4, waic07_5)
mod_comp
```

Overthinking

```{stan output.var="m07_5b"}
data {
  int N;
  vector[N] loggdp;
  vector[N] rugged;
  vector<lower=0,upper=1>[N] cont_africa;
}
parameters {
  real a;
  real bR;
  real bA;
  real bAR;
  real<lower=0, upper=10> sigma;
}
model {
  vector[N] mu;
  
  for (i in 1:N) {
    mu[i] = a + bR * rugged[i] + bAR * rugged[i] * cont_africa[i] + bA * cont_africa[i];
  }

  target += normal_lpdf(loggdp | mu, sigma);
  target += normal_lpdf(a | 8, 100);
  target += normal_lpdf(bR | 0, 1);
  target += normal_lpdf(bA | 0, 1);
  target += normal_lpdf(bAR | 0, 1);
}
generated quantities {
  vector[N] log_lik;
  {
  vector[N] mu; vector[N] gamma;
  for(n in 1:N) {
    gamma[n] = bR + bAR * cont_africa[n];
    mu[n] = a + gamma[n] * rugged[n] + bA * cont_africa[n];
    log_lik[n] = normal_lpdf(loggdp[n] | mu[n], sigma);
  }
  }
}

```

### 7.1.3 Plotting the interaction

```{r}
post07_5 <- as.data.frame(fit07_5)

f_mu_07_5 <- function(rugged, cont_africa) with(post07_5, 
             a + bR * rugged + bAR * rugged * cont_africa + bA * cont_africa )

mu_07_5 <- mapply(f_mu_07_5, rugged = d$rugged, cont_africa = d$cont_africa)

mu_07_5_mean <- apply(mu_07_5, 2, mean)
mu_07_5_pi <- apply(mu_07_5, 2, rethinking::PI, prob = .97)

d_07_5 <- d %>%
  mutate(mu_mean = mu_07_5_mean,
         mu_pi_l = mu_07_5_pi[1,],
         mu_pi_h = mu_07_5_pi[2,],
         inAfrica = ifelse(cont_africa, 'African nations', 'Non-African nations'))
```

Figure 7.4

```{r}
ggplot(d_07_5) +
  geom_point(aes(rugged, loggdp, color = cont_africa), shape = 21) +
  theme(legend.position = '') +
  geom_line(aes(rugged, mu_mean)) +
  geom_ribbon(aes(x=rugged, ymin=mu_pi_l, ymax=mu_pi_h, fill = cont_africa), alpha = .1) +
  facet_wrap(~inAfrica) +
  labs(x = 'Terrain Ruggedness Index', y = 'log GDP year 2000')
```

#### 7.1.4.1 parameters change meaning

Model summary

```{r}
print(fit07_5, probs = c(0.1, 0.5, 0.9), pars = c('a', 'bA', 'bR', 'bAR', 'sigma'))
```

#### 7.1.4.2 Incorporating uncertainty

Calculate posterior for gamma

```{r}
post <- as.data.frame(fit07_5, pars = c('a', 'bA', 'bR', 'bAR'))
gamma_africa <- post$bR + post$bAR * 1
gamma_notafrica <- post$bR + post$bAR * 0
```

means

```{r}
c(mean(gamma_africa), mean(gamma_notafrica))
```

Figure 7.5

```{r}
ggplot() +
  geom_density(aes(gamma_africa), color = 'dodgerblue') +
  geom_density(aes(gamma_notafrica), color = 'black') +
  labs(x = 'gamma')
```

Difference between them

```{r}
diff <- gamma_africa - gamma_notafrica
sum( diff < 0 ) / length( diff )
```

## 7.2 Symmetric of the liner interaction

### 7.2.2 Africa depends on ruggedness

```{r}
d_07_5 <- d_07_5 %>%
  mutate(rugged_hi = rugged > median(rugged),
         inAfrica_num = as.integer(inAfrica=='African nations')) %>%
  mutate(inAfrica_num = ifelse(rugged_hi, inAfrica_num + .02, inAfrica_num))

d_imaginary <- d_07_5 %>% select(rugged) %>% range

# imaginary low values
mu_rugged_lo <- mapply(f_mu_07_5, rugged = d_imaginary[1], cont_africa = 0:1)
mu_rugged_lo_mean <- apply(mu_rugged_lo, 2, mean)
mu_rugged_lo_pi <- apply(mu_rugged_lo, 2, rethinking::PI, prob = .97)

# imaginary high values
mu_rugged_hi <- mapply(f_mu_07_5, rugged = d_imaginary[2], cont_africa = 0:1)
mu_rugged_hi_mean <- apply(mu_rugged_hi, 2, mean)
mu_rugged_hi_pi <- apply(mu_rugged_hi, 2, rethinking::PI, prob = .97)

# organize data
d_imag <- 
  data.frame(inAfrica_num = 0:1, 
             country = rep(c('lo', 'hi'), each = 2), 
             loggdp_mean = c(mu_rugged_lo_mean, mu_rugged_hi_mean), 
             loggdp_pi_l = c(mu_rugged_lo_pi[1,], mu_rugged_hi_pi[1,]), 
             loggdp_pi_h = c(mu_rugged_lo_pi[2,], mu_rugged_hi_pi[2,]))

```

Figure 7.6

```{r}
ggplot() +
  # graph raw actual values
  geom_point(data = d_07_5,
             aes(inAfrica_num, loggdp, color = rugged_hi), shape = 21) +
  
  # graph imaginary hi values
  geom_ribbon(data = d_imag %>% filter(country == 'hi'),
              aes(x = inAfrica_num, y = loggdp_mean, ymin = loggdp_pi_l, ymax = loggdp_pi_h),
              alpha = .1, fill = 'dodgerblue') +
  geom_line(data = d_imag %>% filter(country == 'hi'),
              aes(x = inAfrica_num, y = loggdp_mean),
              color = 'dodgerblue') +
  
  # graph imaginary lo values
  geom_ribbon(data = d_imag %>% filter(country == 'lo'),
              aes(x = inAfrica_num, y = loggdp_mean, ymin = loggdp_pi_l, ymax = loggdp_pi_h),
              alpha = .1, fill = 'black') +
  geom_line(data = d_imag %>% filter(country == 'lo'),
              aes(x = inAfrica_num, y = loggdp_mean),
              color = 'black', linetype = 'dashed') +
  
  # other stuff
  scale_color_manual(values = c('black', 'dodgerblue')) +
  theme(legend.position = '') +
  labs(x = 'continent', y = 'log GDP year 2000') +
  scale_x_continuous(breaks = c(0.01, 1.01), labels = c('other', 'Africa'))
```

## 7.3 Continuous interactions

### 7.3.1 The data

Load the data

```{r}
data('tulips', package = 'rethinking')
d  <- tulips; rm(tulips)
```

### 7.3.2 The uncentered model

Code the two models

```{stan output.var="m07_6"}
data {
  int N;
  vector[N] water;
  vector[N] shade;
  vector[N] blooms;
}
parameters {
  real a;
  real bW;
  real bS;
  real<lower=0> sigma;
}
model {
  vector[N] mu;
  target += normal_lpdf(a | 0, 100);
  target += normal_lpdf(bS | 0, 100);
  target += normal_lpdf(bW | 0, 100);
  target += uniform_lpdf(sigma | 0, 100);
  
  mu = a + bW * water + bS * shade;
  target += normal_lpdf(blooms | mu, sigma);
  
}
generated quantities {
  vector[N] log_lik;
  {
  vector[N] mu;
  for(i in 1:N) {
    mu[i] = a + bW * water[i] + bS * shade[i];
    log_lik[i] = normal_lpdf(blooms[i] | mu[i], sigma);
  }
  }
}

```


```{stan output.var="m07_7"}
data {
  int N;
  vector[N] water;
  vector[N] shade;
  vector[N] blooms;
}
parameters {
  real a;
  real bW;
  real bS;
  real bWS;
  real<lower=0> sigma;
}
model {
  vector[N] mu;
  target += normal_lpdf(a | 0, 100);
  target += normal_lpdf(bS | 0, 100);
  target += normal_lpdf(bW | 0, 100);
  target += normal_lpdf(bWS | 0, 100);
  target += uniform_lpdf(sigma | 0, 100);
  
  mu = a + bW * water + bS * shade + bWS * water .* shade;
  target += normal_lpdf(blooms | mu, sigma);
  
}
generated quantities {
  vector[N] log_lik;
  {
  vector[N] mu;
  for(i in 1:N) {
    mu[i] = a + bW * water[i] + bS * shade[i] + bWS * water[i] * shade[i];
    log_lik[i] = normal_lpdf(blooms[i] | mu[i], sigma);
  }
  }
}

```

Organize data, fit, and sample from both models.

```{r}
dat <- list(N = NROW(d), water = d$water, shade = d$shade, blooms = d$blooms)
fit07_6 <- sampling(m07_6, data = dat, iter = 1000, chains = 2, cores = 2)
fit07_7 <- sampling(m07_7, data = dat, iter = 1000, chains = 2, cores = 2)
```

Summary of models

```{r}
print(fit07_6, include = F, pars = 'log_lik', probs = c(0.1, 0.5, 0.9))
print(fit07_7, include = F, pars = 'log_lik', probs = c(0.1, 0.5, 0.9))
```

Compare models

```{r}
log_lik_07_6 <- extract_log_lik(fit07_6, merge_chains = FALSE)
waic07_6 <- waic(log_lik_07_6)

log_lik_07_7 <- extract_log_lik(fit07_7, merge_chains = FALSE)
waic07_7 <- waic(log_lik_07_7)

mod_comp <- compare(waic07_6, waic07_7)
mod_comp
```

For centering, we can use the same models, and just center the data given to them.

```{r}
dat <- list(N = NROW(d), water = d$water - mean(d$water), shade = d$shade - mean(d$shade), blooms = d$blooms)
fit07_8 <- sampling(m07_6, data = dat, iter = 10000, chains = 2, cores = 2)
fit07_9 <- sampling(m07_7, data = dat, iter = 10000, chains = 2, cores = 2)
```

Now summarise the centered versions.

```{r}
print(fit07_8, include = F, pars = 'log_lik', probs = c(0.1, 0.5, 0.9))
print(fit07_9, include = F, pars = 'log_lik', probs = c(0.1, 0.5, 0.9))
```

### 7.3.4 Plotting implied predictions

```{r}
# setup new data
nd <- expand.grid(water = seq(-1, 1), shade = seq(-1, 1))

# get draws
post07_8 <- as.data.frame(fit07_8, pars = c('a', 'bW', 'bS'))
post07_9 <- as.data.frame(fit07_9, pars = c('a', 'bW', 'bS', 'bWS'))

# setup posterior calcs
f_mu_07_8 <- function(water, shade) with(post07_8, 
             a + bW * water + bS * shade  )
mu_07_8 <- mapply(f_mu_07_8, water = nd$water, shade = nd$shade)

f_mu_07_9 <- function(water, shade) with(post07_9, 
             a + bW * water + bS * shade + bWS * water * shade )
mu_07_9 <- mapply(f_mu_07_9, water = nd$water, shade = nd$shade)

# summarise posterior
mu_07_8_mean <- apply(mu_07_8, 2, mean)
mu_07_8_pi <- apply(mu_07_8, 2, rethinking::PI, prob = .97)

mu_07_9_mean <- apply(mu_07_9, 2, mean)
mu_07_9_pi <- apply(mu_07_9, 2, rethinking::PI, prob = .97)

# organize data
nd <- cbind(bWS = 0, nd) %>% rbind(cbind(bWS = 1, nd))

nd <- nd %>%
  mutate(bWS = ifelse(bWS, 'With Interaction', 'Without interaction'),
         water = paste0('water.c = ', water),
         blooms = c(mu_07_8_mean, mu_07_9_mean),
         blooms_pi_l = c(mu_07_8_pi[1,], mu_07_9_pi[1,]),
         blooms_pi_h = c(mu_07_8_pi[2,], mu_07_9_pi[2,]))
```

Figure 7.7 bottom row

```{r}
ggplot(nd) +
  geom_ribbon(aes(x=shade, ymin=blooms_pi_l, ymax=blooms_pi_h), alpha = .1) +
  geom_line(aes(shade, blooms)) +
  scale_x_continuous(breaks = c(-1, 0, 1)) +
  theme(panel.border = element_rect(fill = NA),
        panel.spacing = unit(0, 'mm')) +
  facet_grid(bWS~water)
```

Not sure what was mean by data in the points as we created new data to predict blooms here.






