---
title: "Chapter 14"
author: "Scott Spencer"
date: "9/6/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, message = FALSE, error = FALSE)
library(dplyr); library(tidyr); library(rstan); library(skimr); library(ggplot2); library(ggthemes)
theme_set(theme_tufte(base_family = 'sans'))
```

The code below is meant as a directly-in-Stan translation of the examples in Chapter 14 of McElreath's *Statistical Rethinking*.

## 14.1 Measurement error

load the data.

```{r}
data('WaffleDivorce', package = 'rethinking')
d <- WaffleDivorce; rm(WaffleDivorce)
```

Figure 14.1


```{r}
p1 <- ggplot(d) + theme_tufte(base_family = 'sans') +
  geom_segment(aes(x = MedianAgeMarriage, xend = MedianAgeMarriage, 
                   y = Divorce + Divorce.SE, yend = Divorce - Divorce.SE)) +
  geom_point(aes(MedianAgeMarriage, Divorce), shape = 21, fill = 'white') + 
  theme(plot.margin=unit(c(0,1,0,1),"cm")) +
  labs(x = 'Median age marriage', y = 'Divorce rate')

p2 <- ggplot(d) + theme_tufte(base_family = 'sans') +
  geom_segment(aes(x = log(Population), xend = log(Population), 
                   y = Divorce + Divorce.SE, yend = Divorce - Divorce.SE)) +
  geom_point(aes(log(Population), Divorce), shape = 21, fill = 'white') + 
  theme(plot.margin=unit(c(0,0,0,1),"cm")) +
  labs(x = 'Log population', y = 'Divorce rate')

library(gridExtra)
grid.arrange(p1, p2, nrow = 1)
```


### 14.1.1 Error on the outcome

Code model in Stan.

```{stan output.var="m14_1"}
data {
  int N;
  vector[N] A;
  vector[N] R;
  vector[N] Dobs;
  vector[N] Dsd;
}
parameters {
  real a;
  real ba;
  real br;
  real<lower=0> sigma;
  vector[N] Dest;
}
model {
  vector[N] mu; 
  // priors
  target += normal_lpdf(a | 0, 10);
  target += normal_lpdf(ba | 0, 10);
  target += normal_lpdf(br | 0, 10);
  target += cauchy_lpdf(sigma | 0, 2.5);
  
  // linear model
  mu = a + ba * A + br * R;
  
  // likelihood
  target += normal_lpdf(Dest | mu, sigma);
  
  // prior for estimates
  target += normal_lpdf(Dobs | Dest, Dsd);
}
generated quantities {
  vector[N] log_lik;
  {
  vector[N] mu;
  mu = a + ba * A + br * R;
  for(i in 1:N) log_lik[i] = normal_lpdf(Dest[i] | mu[i], sigma);
  }
}

```

Organize data and sample from model.

```{r}
dat <- list(
  N = NROW(d),
  A = d$MedianAgeMarriage,
  R = d$Marriage,
  Dobs = d$Divorce,
  Dsd = d$Divorce.SE
)

fit14_1 <- sampling(m14_1, data = dat, iter = 1000, chains = 2, cores = 2)
```

Figure 14.2 left side

Show effects of shrinkage from the model.

```{r}
Dest14_1 <- as.matrix(fit14_1, pars = 'Dest')
d <- d %>% mutate(Dest_mean = apply(Dest14_1, 2, mean),
                  Dest_sd = apply(Dest14_1, 2, sd))

ggplot(d) + theme_tufte(base_family = 'sans') +
  geom_point(aes(x = Divorce.SE, y = Dest_mean - Divorce)) +
  geom_hline(yintercept = 0, linetype = 'dashed') +
  labs(x = 'Divorce observed standard error', 
       y = 'Divorce estimated - divorce observed')
```

Figure 14.2 right side

```{r}
ggplot(d) + theme_tufte(base_family = 'sans') +
  geom_segment(aes(x = MedianAgeMarriage, xend = MedianAgeMarriage, 
                   y = Dest_mean - Dest_sd, yend = Dest_mean + Dest_sd)) +
  geom_point(aes(x = MedianAgeMarriage, y = Dest_mean), shape = 21, fill = 'white') +
  geom_smooth(aes(x = MedianAgeMarriage, y = Divorce), method='lm',formula=y~x, alpha = .1, linetype = 'dashed', color = 'black', lwd = .5) +
  geom_smooth(aes(x = MedianAgeMarriage, y = Dest_mean), method='lm',formula=y~x, alpha = .2, lwd = .5)
# TODO: THIS DOESN'T MATCH BOOK
```

### 14.1.2 Error on both outcome and predictor

Code model in Stan.

```{stan output.var="m14_2"}
data {
  int N;
  vector[N] A;
  vector[N] Dobs;
  vector[N] Dsd;
  vector[N] Robs;
  vector[N] Rsd;
}
parameters {
  real a;
  real ba;
  real br;
  real<lower=0> sigma;
  vector[N] Dest;
  vector[N] Rest;
}
model {
  vector[N] mu; 
  // priors
  target += normal_lpdf(a | 0, 10);
  target += normal_lpdf(ba | 0, 10);
  target += normal_lpdf(br | 0, 10);
  target += cauchy_lpdf(sigma | 0, 2.5);
  
  // linear model
  mu = a + ba * A + br * Rest;
  
  // likelihood
  target += normal_lpdf(Dest | mu, sigma);
  
  // prior for estimates
  target += normal_lpdf(Dobs | Dest, Dsd);
  target += normal_lpdf(Robs | Rest, Rsd);
}
generated quantities {
  vector[N] log_lik;
  {
  vector[N] mu;
  mu = a + ba * A + br * Rest;
  for(i in 1:N) log_lik[i] = normal_lpdf(Dest[i] | mu[i], sigma);
  }
}

```

Organize data and sample from model.

```{r}
dat <- list(
  N = NROW(d),
  A = d$MedianAgeMarriage,
  Robs = d$Marriage,
  Rsd = d$Marriage.SE,
  Dobs = d$Divorce,
  Dsd = d$Divorce.SE
)

fit14_2 <- sampling(m14_2, data = dat, iter = 1000, chains = 2, cores = 2)
```

Summarise the model

```{r}
print(fit14_2)
```

Figure 14.3 left side

```{r}
Rest14_2 <- as.matrix(fit14_2, pars = 'Rest')
d <- d %>% mutate(Rest_mean = apply(Rest14_2, 2, mean),
                  Rest_sd = apply(Rest14_2, 2, sd))

ggplot(d) + theme_tufte(base_family = 'sans') +
  geom_point(aes(x = Marriage.SE, y = Rest_mean - Marriage)) +
  geom_hline(yintercept = 0, linetype = 'dashed') +
  labs(x = 'Marriage observed standard error', 
       y = 'Marriage estimated - Marriage observed')
```

Figure 14.3 right side
```{r}
ggplot(d) + theme_tufte(base_family = 'sans') +
  geom_segment(aes(x = Marriage, xend = Rest_mean,
                   y = Divorce, yend = Dest_mean)) +
  geom_point(aes(x = Marriage, y = Divorce), color = 'dodgerblue') +
  geom_point(aes(x = Rest_mean, y = Dest_mean), shape = 21, fill = 'white') +
  labs(x = 'Marriage rate (posterior)', y = 'Divorce rate (posterior)')

```

## 14.2 Missing data

### 14.2.1 Imputing neocortex

Load the data.

```{r}
data('milk', package = 'rethinking')
d <- milk; rm(milk)
d <- d %>% 
  mutate(neocortex.prop = neocortex.perc / 100,
         logMass = log(mass),
         miss_idx = is.na(neocortex.perc) * cumsum(is.na(neocortex.perc)) )
```

Code the MCAR model in Stan.

```{stan output.var="m14_3"}
data {
  int N;
  vector[N] NCobs;
  int N_miss;
  int<lower=0,upper=N_miss> miss_idx[N];
  vector[N] logMass;
  vector[N] k;
}
parameters {
  real a;
  real bn;
  real bm;
  real<lower=0> sigma;
  real mu_nc;
  real<lower=0> sigma_nc;
  vector[N_miss] nc_impute;
}
model {
  vector[N] mu;
  vector[N] NC;
  int j = 1;
  
  // priors
  target += normal_lpdf(a  | 0, 100);
  target += normal_lpdf(bn | 0, 10);
  target += normal_lpdf(bm | 0, 10);
  target += cauchy_lpdf(sigma | 0, 1);
  target += normal_lpdf(mu_nc | 0.5, 1);
  target += cauchy_lpdf(sigma_nc | 0, 1);
  
  // combine observed and estimates for missing
  NC = NCobs;
  for(i in 1:N) if(miss_idx[i] > 0) NC[i] = nc_impute[miss_idx[i]];

  // impute missing
  target += normal_lpdf(NC | mu_nc, sigma_nc);
  
  // linear model
  mu = a + bn * NC + bm * logMass;
  
  // likelihood
  target += normal_lpdf(k | mu, sigma);
}

```

Organize data and sample from the model.

```{r}
dat <- list(
  N = NROW(d),
  NCobs = ifelse(is.na(d$neocortex.prop), -1, d$neocortex.prop),
  N_miss = sum(is.na(d$neocortex.perc)),
  miss_idx = d$miss_idx,
  logMass = d$logMass,
  k = d$kcal.per.g
)

fit14_3 <- sampling(m14_3, data = dat, iter = 10000, chains = 2, cores = 2)
```

Summarise model.

```{r}
print(fit14_3, probs = c(.1, .5, .9))
```

Compare above model to one using only complete cases.

```{stan output.var="m14_4"}
data {
  int N;
  vector[N] NC;
  vector[N] logMass;
  vector[N] k;
}
parameters {
  real a;
  real bn;
  real bm;
  real<lower=0> sigma;
}
model {
  vector[N] mu;

  // priors
  target += normal_lpdf(a  | 0, 100);
  target += normal_lpdf(bn | 0, 10);
  target += normal_lpdf(bm | 0, 10);
  target += cauchy_lpdf(sigma | 0, 1);

  // linear model
  mu = a + bn * NC + bm * logMass;
  
  // likelihood
  target += normal_lpdf(k | mu, sigma);
}

```

Organize data and sample from model

```{r}
dcc <- filter(d, complete.cases(d))

dat <- list(
  N = NROW(dcc),
  NC = dcc$neocortex.prop,
  logMass = dcc$logMass,
  k = dcc$kcal.per.g
)

fit14_4 <- sampling(m14_4, data = dat, iter = 10000, chains = 2, cores = 2)

```

Summarise model

```{r}
print(fit14_4, probs = c(.1, .5, .9))
```

Figure 14.4 left side

```{r}

imputed14_3 <- 
  as.data.frame(fit14_3, pars = c('nc_impute')) %>%
  gather %>% 
  group_by(key) %>%
  summarise(nc_imp_mean = mean(value),
            nc_pi_l = rethinking::PI(value)[1],
            nc_pi_h = rethinking::PI(value)[2]) %>%
  ungroup() %>%
  mutate(obs = as.integer(gsub('[^0-9]', '', key))) %>% 
  arrange(obs) %>%
  mutate(kcal.per.g = d[!complete.cases(d),]$kcal.per.g,
         logMass = d[!complete.cases(d),]$logMass)

# d$obsest <- d$neocortex.prop
# d[!complete.cases(d),]$obsest <- imputed14_3$nc_imp_mean

# TODO: regression line doesn't match book

ggplot() + 
  # stat_smooth(data = d,
  #             aes(x = obsest, y = kcal.per.g),
  #             method = lm, alpha = .2) +
  geom_segment(data = imputed14_3,
               aes(x = nc_pi_l, xend = nc_pi_h,
                   y = kcal.per.g, yend = kcal.per.g)) +
  geom_point(data = imputed14_3,
             aes(x = nc_imp_mean, y = kcal.per.g), shape = 21, fill = 'white') +
  geom_point(data = d,
             aes(x = neocortex.prop,
                 y = kcal.per.g), color = 'dodgerblue') +
  labs(x = 'neocortex proportion', y = 'kcal per gram')
```

Figure 14.4 right side

```{r}
ggplot() + 
  geom_point(data = d,
             aes(x = logMass, y = neocortex.prop), color = 'dodgerblue') +
  geom_segment(data = imputed14_3,
               aes(x = logMass, xend = logMass, 
                   y = nc_pi_l, yend = nc_pi_h)) +
  geom_point(data = imputed14_3,
             aes(x = logMass, y = nc_imp_mean), shape = 21, fill = 'white') +
  scale_x_continuous(breaks = seq(-2, 4))
```

### 14.2.2 Improving the imputation model

Code improved model in Stan.

```{stan output.var="m14_5"}
data {
  int N;
  vector[N] nc_obs;
  int N_missing;
  int<lower=0, upper=N_missing> missing[N];
  vector[N] k;
  vector[N] logmass;
}
parameters {
  vector[N_missing] nc_imp;
  real a;
  real bn;
  real bm;
  real gm;
  real a_nc;
  real<lower=0> sigma_nc;
  real<lower=0> sigma;
}
model {
  vector[N] mu;
  vector[N] mu_nc;
  vector[N] nc;
  
  // priors
  target += normal_lpdf(a | 0, 100);
  target += normal_lpdf(bn | 0, 10);
  target += normal_lpdf(bm | 0, 10);
  target += normal_lpdf(gm| 0, 10);
  target += normal_lpdf(a_nc | 0.5 , 1);
  target += cauchy_lpdf(sigma_nc | 0, 1);
  target += cauchy_lpdf(sigma | 0, 1);

  // imputation
  nc = nc_obs;
  for(i in 1:N) if (missing[i] > 0) nc[i] = nc_imp[missing[i]];
  
  mu_nc = a_nc + gm * logmass;
  target += normal_lpdf(nc | mu_nc, sigma_nc);
  
  // linear model
  mu = a + bn * nc + bm * logmass;

  // likelihood
  target += normal_lpdf(k | mu, sigma);
}

```

Organize data and sample from the model.

```{r}
dat <- list(
  N = NROW(d),
  nc_obs = ifelse(is.na(d$neocortex.prop), -1, d$neocortex.prop),
  N_missing = sum(is.na(d$neocortex.perc)),
  missing = d$miss_idx,
  logmass = d$logMass,
  k = d$kcal.per.g
)

fit14_5 <- sampling(m14_5, data = dat, iter = 10000, chains = 2, cores = 2)
```

Summarise model.

```{r}
print(fit14_5, probs = c(.1, .5, .9))
```

Figure 14.5 left side

```{r}

imputed14_5 <- 
  as.data.frame(fit14_5, pars = c('nc_imp')) %>%
  gather %>% 
  group_by(key) %>%
  summarise(nc_imp_mean = mean(value),
            nc_pi_l = rethinking::PI(value)[1],
            nc_pi_h = rethinking::PI(value)[2]) %>%
  ungroup() %>%
  mutate(obs = as.integer(gsub('[^0-9]', '', key))) %>% 
  arrange(obs) %>%
  mutate(kcal.per.g = d[!complete.cases(d),]$kcal.per.g,
         logMass = d[!complete.cases(d),]$logMass)

# TODO: regression line doesn't match book

ggplot() + theme_tufte(base_family = 'sans') +
  # stat_smooth(data = d, 
  #             aes(x = neocortex.prop, y = kcal.per.g),
  #             method = lm, alpha = .2) +
  geom_segment(data = imputed14_5,
               aes(x = nc_pi_l, xend = nc_pi_h,
                   y = kcal.per.g, yend = kcal.per.g)) +
  geom_point(data = imputed14_5,
             aes(x = nc_imp_mean, y = kcal.per.g), shape = 21, fill = 'white') +
  geom_point(data = d,
             aes(x = neocortex.prop,
                 y = kcal.per.g), color = 'dodgerblue') +
  labs(x = 'neocortex proportion', y = 'kcal per gram')
```

Figure 14.5 right side

```{r}
ggplot() + theme_tufte(base_family = 'sans') +
  geom_point(data = d,
             aes(x = logMass, y = neocortex.prop), color = 'dodgerblue') +
  geom_segment(data = imputed14_5,
               aes(x = logMass, xend = logMass, 
                   y = nc_pi_l, yend = nc_pi_h)) +
  geom_point(data = imputed14_5,
             aes(x = logMass, y = nc_imp_mean), shape = 21, fill = 'white') +
  scale_x_continuous(breaks = seq(-2, 4))
```

